% UC Merced  PhD Dissertation Template
% (UCSD Mathematics Dissertation Template is modified according to the UC Merced guidelines by Lasith Adhikari. Not responsible for any issues regarding modifications. You are welcome to add/update if there is any missing requirernment)
%
% Please read the comments in this file and make appropriate edits.
% NOTE: Always refer to the ``Preperation and Submission Manual for 
% Doctoral Dissertations and Masters Theses for 20**'', where 20** is 
% the year of your graduation, for officiation preparations guidelines.
%
% If you desire more control, please see the attached files:
%   * ucsd.cls -- Class file
%   * uct10.clo, uct11.clo, uct12.clo -- Configuration files for font sizes 10pt,11pt,12pt
%
% CHANGELOG:
%   * Original file adapted from brockman.tex by JRB and RMR
%     to work with ucsd.cls


\documentclass[12pt,chapterheads]{UCMerced}
% documentclass options: default is 11pt, oneside, final.
% fonts: 10pt, 11pt, 12pt -- are valid for UCSD dissertations (now UC Merced).
% sides: oneside, twoside -- note that two-sided theses are not accepted by OGS
% mode: draft, final -- draft mode switches to single spacing, removes hyperlinks,
%                       and places a black box at every overfull hbox (check these before submission).
% chapterheads -- include this if you want your chapters to read:
% Chapter 1
% Title of Chapter
%
% instead of
%
% 1 Title of Chapter


% Include all packages you need here.  Some standard options are suggested below.

% GEOMETRY - This will force the use of Letter paper.
% Many TeX installations default to A4 paper.  The formatting
% of the thesis class file requires Letter, else the margins
% will be wrong when you go to print it (and OGS will complain).
% If your TeX implementation is not setup for Letter paper, and
% you cannot change it, uncommenting the following line may fix 
% problem.
% \usepackage[paper=letterpaper]{geometry}

\usepackage{graphicx}

%% AMS PACKAGES - Chances are you will want some or all of these if writing a math dissertation.
\usepackage{amsmath, amscd, amssymb, amsthm}
% \usepackage{amssymb,amsmath}
\usepackage{multicol}
\usepackage{tabulary}
\usepackage{tabularx}
\usepackage{pgfplots}
\usepackage{longtable}
% \usepackage{url}
\usepackage{float}
\usepackage{subcaption}
\usepackage{booktabs}

\usepackage{gb4e}  % linguistic examples


%% LATIN MODERN FONTS (replacements for Computer Modern)
\usepackage{lmodern}
\usepackage[T1]{fontenc}
\usepackage{adjustbox}

%% INDEX
% Uncomment the following two lines to create an index: 
% \usepackage{makeidx}
% \makeindex
% You will need to uncomment the \printindex line near the
% bibliography to display the index.  Use the command
% \index{keyword} within the text to create an entry in the index
% for keyword.

%% HYPERLINKS
% To create a PDF with hyperlinks, you need to include the hyperref package.
% THIS HAS TO BE THE LAST PACKAGE INCLUDED!
% Note that the options plainpages=false and pdfpagelabels exist
% to fix indexing associated with having both (ii) and (2) as pages.
% Also, all links must be black according to OGS.
% See: http://www.tex.ac.uk/cgi-bin/texfaq2html?label=hyperdupdest
% Note: This may not work correctly with all DVI viewers (i.e. Yap breaks).
\usepackage[colorlinks=true, pdfstartview=FitV, linkcolor=black, citecolor=black, urlcolor=black,plainpages=false,pdfpagelabels]{hyperref}
\hypersetup{ pdfauthor = {Your Name Here}, pdftitle = {The Title of The Dissertation}, pdfkeywords = {Keywords for Searching}, pdfcreator = {pdfLaTeX with hyperref package}, pdfproducer = {pdfLaTeX}}

\usepackage{cleveref}
%% GRAPHICX - This is the standard package for including graphics for latex/pdflatex.
\usepackage{apacite}

\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\obspre}{$\langle o_{pre} \rangle$}
\DeclareMathOperator*{\obspost}{$\langle o_{post} \rangle$}

\usepackage{url}
\noautomath
\begin{document}

%% REQUIRED FIELDS -- Replace with the values appropriate to you
\title{Four studies of communicative, cognitive, and social factors in extremism and polarization}
% No symbols, formulas, superscripts, or Greek letters are allowed
% in your title.

\author{Matthew A. Turner}
\degreeyear{2021}
\degree{Doctor of Philosophy} 
% Master's Degree theses will NOT be formatted properly with this
% file.

\field{Cognitive and Information Sciences}

\numberofmembers{4} % |chair|  + |othermembers| (do not count co-chair) % change here

\chair{Paul E. Smaldino}
% Uncomment the next line iff you have a Co-Chair
%\cochair{Professor Cochair Semimaster} 

\memberone{Teenie Matlock}
\membertwo{Christopher T. Kello}
\memberthree{Jeffrey Yoshimi}


\begin{frontmatter}
\makefrontmatter % The title, copyright, and signature pages.

%% DEDICATION
% You have three choices here:
%   1. Use the ``dedication'' environment.   Put in the text you want,
%   and you'll get a perfectly respectable dedication page.
%
%   2. Use the ``mydedication'' environment.  If you don't like the
%   formatting of option 1, use this environment and format things
%   however you wish.
%
%   3. If you don't want a dedication, it's not required.


\begin{dedication} % The style file will format this for you.
  To Phoebe Ruth Bernacchi-Turner.  \\[2em]
  As Daniel Tiger and his parents sing, ``We gotta look a little 
  closer to see just how things go.'' Taking care of you makes me happy, 
  too---and caring for you has pushed me to try to 
  understand what is going on between people 
  in this crazy world, which
  will maybe contribute a little bit to making the world better for you and
  any future family members. Thank you for getting me out of my own
  head and back into thinking about what I can do to make the world a better
  place. Maybe some day this will inspire you in your own pursuits of new skills and
  a deeper understanding of the world.
\end{dedication}

% \begin{mydedication} % You are responsible for formatting here.
%   \vspace{1in}
%   \begin{flushleft}
% 	To me.
%   \end{flushleft}
%   
%   \vspace{2in}
%   \begin{center}
% 	And you.
%   \end{center}
% 
%   \vspace{2in}
%   \begin{flushright}
% 	Which equals us.
%   \end{flushright}
% \end{mydedication}


%% EPIGRAPH
%  The same choices that applied to the dedication apply here.

\begin{epigraph} % The style file will position the text for you.
  \emph{Toute v\'{e}rit\'{e} et toute action \\
  impliquent un milieu et une subjectivit\'{e} humaine.}\\[2em]
  Every truth and every action \\
  emerges in the context of human culture and human subjectivity.\\[2em]

  ---Jean Paul Sartre, ``L'Existentialisme est un Humanisme'', 1946
\end{epigraph}

% \begin{myepigraph} % You position the text yourself.
%   \vfil
%   \begin{center}
%     {\bf Think! It ain't illegal yet.}
% 
% 	\emph{---George Clinton}
%   \end{center}
% \end{myepigraph}

\tableofcontents
\listoffigures  % Uncomment if you have any figures
\listoftables   % Uncomment if you have any tables


%% ACKNOWLEDGEMENTS
%  While technically optional, you probably have someone to thank.
%  Also, a paragraph acknowledging all coauthors and publishers (if
%  you have any) is required in the acknowledgements page and as the
%  last paragraph of text at the end of each respective chapter. See
%  the OGS Formatting Manual for more information.

\begin{acknowledgements} 
First, I thank my parents for providing me with many rich experiences throughout
childhood that set me on the path to being a scientist,
and for their generous help and encouragement while completing this 
PhD program.

I sincerely thank Paul Smaldino for pushing me to improve with 
more patience than could be reasonably expected of anyone. 
I entered this program unaware of my poor
sense of career direction and lack of discipline.
I thought I could find success freewheeling across
many different subdisciplines. Paul patiently peeled back the layers of
insulation over my tender ego to help me see that I had a lot of work to do
to improve to be the expert I wanted to be. At the same time Paul has
made it clear that he believes I have valuable skills that can and
should be developed further, and has helped me along immesnely with that.

My committe members have all
endured my overly grand first draft ideas and rantings and patiently steered me
towards more practical steps that helped me make real progress on 
tractable pieces of the grand ideas. Teenie Matlock (along with Paul Maglio) first 
took me on when I began the program. I appreciate Teenie's wisdom,
encouragement, and tutelage on metaphor and language, and research in general. 
Her guidance has enabled me to hear and read political discourse in a vastly more
enlightened way, even enabling me to study it scientifically, which I did not
know people did before meeting her. 
I thank Chris Kello for his support, inspiration, and encouragement 
for my grand ideas about timescales in cognitive science even bringing me in
to work with him on a computational project on the subject. I also 
vividly remember Chris giving me a thumbs up from down the hall after the 
Monday brownbag in support of my first modeling results in what became a 
journal article~\cite{Turner2018} and Study 4. I thank Jeff Yoshimi 
for two excellent courses 
(COGS 202 and COGS 269: \emph{Phenomenology})
and for generously tutoring me in the philosophy of science, 
which has been a powerful addition to my theoretical toolkit.

Finally, I thank everyone in the CIS community. These past five years have been a 
stimulating, challenging, and fun experience
that enabled me to learn about and join in the collective project of
cognitive science.
\end{acknowledgements}


%% VITA
%  A brief vita is required in a doctoral thesis. See the OGS
%  Formatting Manual for more information.
\begin{vitapage}
\begin{vita}
\item[2008] B.~S. in Mathematics \& Physics, Syracuse University, Syracuse, New York
\item[2012] M.~S. with Thesis in Applied Physics, Rice University, Houston, Texas
\item[2012-2014] Data Engineer, Economic Modeling Specialists, Int'l, Moscow, Idaho
\item[2014-2016] Research Software Developer II, Northwest Knowledge Network, University of Idaho, Moscow, Idaho
  \item[2016-2020] Graduate Teaching Assistant, University of California, Merced
  \item[2020-2021] Graduate Student Researcher,  University of California, Merced
  \item[2021] Ph.~D. in Cognitive and Information Sciences, University of California, Merced
\end{vita}
\begin{publications}
  \item 
    Turner, M.A., and Smaldino, P.E. (2021). Most group polarization results may be simple conformity. In prep.
  \item
  Turner, M. A., and Smaldino, P. E. (2021). Mechanistic Modeling for the Masses - commentary on Yarkoni, “The generalizability crisis.” \emph{Behavioral and Brain Sciences}, Forthcoming. Retrieved from https://psyarxiv.com/8pj9n.
  \item
    Smaldino, P.E., and Turner, M.A. (2021). 
    Covert signaling is an adaptive communication strategy in diverse populations.
  Under review. Preprint: https://osf.io/preprints/socarxiv/j9wyn/.
  \item 
    Turner, M.A., and Smaldino, P.E. (2020).  Stubborn extremism as a potential pathway to group polarization. 
    In \emph{Proceedings of the 42nd Annual Conference of the Cognitive Science Society}. Online.
  \item
    Smaldino, P.E., Turner, M.A., and Contreras Kallens, P. (2019). Open science and modified funding
    lotteries can impede the natural selection of bad science. \emph{Royal Society Open Science}.
  \item 
    Turner, M.A., and Smaldino, P.E. (2018).  Paths to polarization: how extreme views,
    miscommunication, and random chance drive opinion dynamics.  \emph{Complexity}.
  \item 
    Turner, M.A., Maglio, P.P., and Matlock, T. (2018).  Metaphorical violence
    in political discourse. 
    Under revision. Preprint at \url{https://osf.io/preprints/socarxiv/t8yg9/}.
\end{publications}
\end{vitapage}

%% Abstract
% There does not seem to be a maximum length. From the OGS Formatting
% Manual: ``The abstract may continue on to a second page.''

\begin{abstract}
  Rising extremism and polarization threaten democratic institutions worldwide.
  As opposing factions become more extreme in their opinions, polarization
  widens the chasm between fellow citizens, and common ground erodes, washed
  away down a river of vitriol, bitterness, and hate. What causes increased
  extremism and polarization? Due to the highly complex nature of human
  societies, this problem of explaining polarization must be broken down into
  many sub-problems, which themselves require complex systems thinking to
  address. Simplified models of social systems and rigorous analysis of empirical data,
  are necessary to build a thorough, coherent understanding of social behavior.

  In this dissertation I present my findings from studying three
  sub-problems in explaining why and how extremism and polarization emerge.
  First, I focus narrowly on a communication strategy shown in behavioral
  studies to increase extremism, \emph{metaphorical violence}, such as 
  ``Biden hit Trump over his tax returns in yesterday's debate.'' While we know
  the effects of violence metaphors, we do not understand their distribution
  in the wild, or what causes their usage to increase and decrease. I found
  that metaphorical violence use increased around the time of presidential
  debates and elections in the United States, and was correlated with 
  presidentical candidates' tweets. 

  Second, I show that rising extremism
  in isolated social gropus may be simply explained by the fact that
  extremists are more stubborn than centrists---however existing data on the
  subject is unavailable and behavioral studies on the subject may
  contain ubiquitous false detections of rising extremism. 

  Finally, I developed
  and analyzed an empirically motivated, network theoretic, agent-based model of 
  social influence at the societal level to understand how well we can 
  predict polarization based on the effects of initial conditions, network structure,
  communication noise, and random chance on predictions of polarization.

  Taken together these studies advance our understanding of communicative,
  cognitive, and social factors in the emergence of extremism and polarization.
\end{abstract}
\end{frontmatter}


%% DISSERTATION

% A common strategy here is to include files for each of the chapters. I.e.,
%   \include{chapter1.tex}
%   \include{chapter2.tex}
% etc.  Of course, if you prefer, you can just start with
%   \chapter{My First Chapter Name}
% and start typing away.  
\chapter{Introduction}

\input{Chapter1.tex}

\chapter{Metaphorical violence in political discourse on US cable TV news}

\input{Chapter2.tex}

\chapter{Stubborn extremism explains and predicts group polarization}

\input{Chapter3.tex}

\chapter{Most group polarization results may be simple conformity}


\input{Chapter4.tex}

\chapter{Paths to polarization: extreme views, miscommunication, and random chance}

\input{Chapter5.tex}


% \chapter{}
% \chapter{Title}
% This is only a test.
% \section{A section}

% \appendix
% \chapter{Final notes}
%   Remove me in case of abdominal pain.

%% END MATTER
% \printindex %% Uncomment to display the index
% \nocite{}  %% Put any references that you want to include in the bib 
%               but haven't cited in the braces.

\bibliographystyle{apacite}
\bibliography{/Users/mt/workspace/Papers/library.bib,
    /Users/mt/workspace/Papers/Archive/complexity-overleaf-git/ComplexitySpecialIssueForDissertation.bib
  }
\end{document}

